---
title: "Covid analysis about the situation in Finland"
author: "Jouni Tuomisto"
date: "19 2 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(plotly)

```

## Download data

First we download data from THL open data repository.


```{r data}

# https://thl.fi/fi/tilastot-ja-data/aineistot-ja-palvelut/avoin-data/varmistetut-koronatapaukset-suomessa-covid-19-

dat <- read.csv("data/covid_statistics_history.csv") # Historical, mostly weekly data until 2021-02-15
dat$period <- ifelse(grepl("Vuosi",dat$date), "weekly","daily")
dat <- dat[!is.na(dat$value),]

dat2 <- read.csv("http://77.86.191.32/rtools_server/runs/covid_statistics_daily_fi.csv") #daily updates
dat2$period <- "cumulative"

dat <- rbind(dat, dat2)
colnames(dat)[1:3] <- c("place","age","date")
for(i in c("place","age","date","sex","measure")) {
  dat[[i]] <- as.factor(dat[[i]])
}

start <- as.POSIXct("2019-12-29 12:00 EET")

dat$time <- (start + (as.numeric(substr(dat$date,10,10)) * 53 + as.numeric(substr(dat$date,19,20))) * 7*24*3600)
dat$time[dat$period!="weekly"] <- as.POSIXct(paste0(dat$date[dat$period!="weekly"], " 12:00 EET"))

dat$value[dat$value==".."] <- "-1"
dat$value <- as.integer(dat$value)

# Find the highest weekly case value for each place. Note: for only ca. 10 places the peak occurs before August 2020.

tmp <- dat[dat$period=="weekly" & dat$measure=="cases",] %>%
  mutate(value=as.integer(value)) %>%
  group_by(place) %>%
  filter(value == max(value,na.rm=TRUE))
tmp <- tmp[!duplicated(tmp$place),c("place","time")]
colnames(tmp)[2] <- "peak"  
dat <- merge(dat, tmp, all.x=TRUE)
dat$peak <- dat$time - dat$peak

#####################

# This code should be updated. It contains the location hierarchy.
if(FALSE) {
rl <- readLines("https://sampo.thl.fi/pivot/prod/fi/epirapo/covid19case/fact_epirapo_covid19case.dimensions.json")
rl[1] <- "{"
rl[2] <- '"test":['
rl[length(rl)] <- "}"
rl <- jsonlite::fromJSON(rl)

tst <- unlist(rl, recursive=FALSE)
} # ENDIF

############################3

```

## Data analysis and plots

What is the timeline of cases? What is the timeline relative to the peak value in each place?

```{r plots, echo=TRUE}


tmp <- dat[dat$measure == "cases" & grepl("SHP",dat$place) & dat$period=="weekly" , ]
tmp <- tmp[order(tmp$peak),]

# Plot health care district cases along the timeline
plot_ly(data = tmp, x=~time,  y=~value, type="scatter", mode="lines", fillcolor=~place) %>%
  layout(title="Weekly cases by health care district")

# Plot health care district cases relative to the peak event
plot_ly(data = tmp, x=~peak, y=~as.numeric(value), type="scatter", mode="lines", fillcolor=~place) %>%
  layout(title="Weekly cases by health care district relative to the peak")

tmp <- dat[dat$measure=="cases" & !grepl("SHP",dat$place) & dat$period=="weekly",]
tmp <- tmp[order(tmp$peak),]

# Plot municipality cases relative to the peak event
plot_ly(data = tmp, x=~peak, y=~value, type="scatter", mode="lines", fillcolor=~place) %>%
  layout(title="Weekly cases by municipality relative to the peak")

# See how many-fold the decrease is 2 weeks after the peak
tmp <- tmp[!is.na(tmp$peak) & (tmp$peak == 0 | tmp$peak == 14*24*3600) , ]
tmp$value[tmp$peak!=0] <- 1 / tmp$value[tmp$peak!=0]
tmp <- aggregate(tmp["value"], by = tmp[c("place","measure","period")], FUN = prod)
cat("See how many-fold the decrease is 2 weeks after the largest peak in each place\n
    If a place is not in the list, both values are below the detection limit.\n
    Negative values mean that the peak had that number of cases and then decreased to below the detection limit.\n
    Vantaa is an artifact because the peak is 2021-02-14 and therefore the second value is missing.\n")
tmp[tmp$value!=1,]

muni <- dat[dat$measure=="cases" & dat$period!="daily" & dat$age=="Kaikki ikäryhmät" & dat$sex=="Kaikki sukupuolet",
            !colnames(dat) %in% c("age","sex","measure")]
muni$daily <- muni$value / 7

tmp2 <- muni[muni$period=="cumulative",] # tmp[tmp$date=="2021-02-15",] # Start time of daily follow-up
tmp2$old <- tmp2$value
tmp2$time <- tmp2$time + 24*3600
tmp2 <- tmp2[colnames(tmp2) %in% c("place","time","old")]
muni <- merge(muni, tmp2, all.x=TRUE)
muni$daily <- ifelse(muni$period=="cumulative", muni$value - muni$old, muni$daily)
muni <- muni[!is.na(muni$daily) & muni$time < Sys.time() , ]
muni <- muni[order(muni$time),]
plot_ly(data = muni, x = ~time, y=~daily, fillcolor=~place, type="scatter", mode="lines") %>%
  layout(title="Daily cases of covid-19 by place")


```

## Conclusions

* Most municipalities are able to reduce the number of cases by more than a half in two weeks.
* Important exceptions are Helsinki, Vantaa, Espoo, Rauma, and several small municipalities in Uusimaa.
* After a long, slow worsening, the situation in Uusimaa is quickly starting to get bad since 2021-02-15.
